---
title: "Computer Modeling (OCN 682)"
subtitle: "Intro to Maps Part 1"
author: "Dr. Nyssa Silbiger"
institute: "UH Fall 2026"
date: today
resources:
  - custom-theme.scss
format:
  revealjs:
    theme: [default, custom-theme.scss]
    highlight-style: github
    code-line-numbers: true
    width: 1500
    height: 1000
    preview-links: auto
    self-contained-math: true
    embed-resources: false
    slide-number: c/t
    footer: "OCN 682 | Week 7 | Intro to Maps Part 1"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)
```

## Outline 

Working with Maps

::: {.incremental}
1. Quiz!
1. Overview of spatial packages in R
1. Simple mapping with the [{maps}]{.fn} package + ggplot
1. Preview: the modern [{sf}]{.fn} approach


**Homework:** Make a map using Quarto

::: {.callout-note}
I am NOT a GIS person. Please refer to references on spatial analysis for theory.  
Today is about getting a map on screen with real data.
:::
:::
---

## Show Me Your Quarto Files!

---

## Map Packages in R {.smaller-title}

There are **many** map packages in R. Here is a curated overview:

| Package | What it does | Needs API key? |
|---------|-------------|---------------|
| [{maps}]{.fn} | Simple polygon basemaps (country, state, county) | No |
| [{sf}]{.fn} | Full-featured spatial data (GIS-grade) | No |
| [{rnaturalearth}]{.fn} | World/country/state polygons as sf objects | No |
| [{maptiles}]{.fn} | Download satellite/street tile basemaps | No |
| [{leaflet}]{.fn} | Interactive HTML maps | No |
| [{tmap}]{.fn} | Thematic maps, static and interactive | No |
| [{ggmap}]{.fn} | Tile maps via Google/Stadia | Yes (credit card)~~ |

::: {.callout-tip}
**Today:** [{maps}]{.fn} for polygons and a preview of [{sf}]{.fn}.  
**Next class:** [{maptiles}]{.fn} + [{leaflet}]{.fn} for tile and interactive maps — no API key required!
:::

**Helpful references:**
- [Geocomputation with R](https://geocompr.robinlovelace.net/adv-map.html)
- [Bathymetry example](http://ben-williams.github.io/updated_ggplot_figures.html#maps)
- [Earth Lab — CRS types](https://www.earthdatascience.org/courses/earth-analytics/spatial-data-r/understand-epsg-wkt-and-other-crs-definition-file-types/)

---

## Start Your Script or Quarto File

Install today's packages if you haven't already:

```{r}
#| echo: true
#| eval: false

install.packages(c("maps", "mapdata", "mapproj"))
```

Load libraries:

```{r}
#| echo: true
#| message: false
#| warning: false

library(tidyverse)
library(here)
library(maps)
library(mapdata)
library(mapproj)
```

---

## Today's Data

```{r}
#| echo: true
#| message: false
#| warning: false

# Population in California by county
popdata <- read_csv(here("Week_07", "data", "CApopdata.csv"))

# Seastar counts at field sites in California
stars <- read_csv(here("Week_07", "data", "stars.csv"))
```

**Take a minute to explore both datasets before moving on.**

---

## The {maps} Package {.inverse}

[map_data()]{.fn} pulls polygon boundaries for countries, states, and counties — ready to use with ggplot.

---

## map_data() — World

```{r}
#| echo: true
#| code-line-numbers: "1"

world <- map_data("world")

head(world)
```

---

## map_data() — USA

```{r}
#| echo: true
#| code-line-numbers: "1"

usa <- map_data("usa")

head(usa)
```

---

## map_data() — A Country

```{r}
#| echo: true
#| code-line-numbers: "1"

italy <- map_data("italy")

head(italy)
```

---

## map_data() — US States

```{r}
#| echo: true
#| code-line-numbers: "1"

states <- map_data("state")

head(states)
```

---

## map_data() — US Counties

```{r}
#| echo: true
#| code-line-numbers: "1"

counties <- map_data("county")

head(counties)
```

---

## Structure of the Data

::: {.incremental}
- **[long]{.var}** — longitude. West of the prime meridian is negative.
- **[lat]{.var}** — latitude.
- **[order]{.var}** — the order in which ggplot connects the dots (polygon vertices).
- **[region]{.var} / [subregion]{.var}** — the region or subregion a set of points surrounds.
- **[group]{.var}** — *This one is critical!* Controls whether adjacent points are connected by lines. Points in the same group get connected; different groups cause ggplot to "lift the pen."


::: {.callout-warning}
Forgetting [`group = group`]{.param} in your aesthetic is the #1 cause of "crazy line" plots. You will do this. Everyone does.
:::
:::
---

## Make a Map of the World

```{r}
#| echo: true
#| output-location: column
#| fig-width: 7
#| fig-height: 5

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat,
                   group = group))
```

---

## What Happens Without `group = group`?

```{r}
#| echo: true
#| output-location: column
#| code-line-numbers: "3"
#| fig-width: 7
#| fig-height: 5

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat))
```

::: {.callout-warning}
Those "crazy lines" mean ggplot is connecting all points in one continuous path — it has no idea where one polygon ends and the next begins.
:::

---

## Add Color to Lines and Fill

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot1
#| echo: true
#| eval: false
#| code-line-numbers: "5"

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat,
                   group = group),
               color = "black")
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot1
#| echo: false
#| warning: false
#| fig-width: 7
#| fig-height: 5
```
:::
:::

---

## Fill by Region

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot1a
#| echo: true
#| eval: false
#| code-line-numbers: "5"

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat,
                   group = group,
                   fill = region),
               color = "black") +
  guides(fill = "none")
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot1a
#| echo: false
#| warning: false
#| fig-width: 7
#| fig-height: 5
```
:::
:::

---

## Change the Theme

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot2
#| echo: true
#| eval: false
#| code-line-numbers: "8"

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat,
                   group = group,
                   fill = region),
               color = "black") +
  guides(fill = "none") +
  theme_minimal()
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot2
#| echo: false
#| warning: false
#| fig-width: 7
#| fig-height: 5
```
:::
:::

---

## Make the Ocean Blue

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot3
#| echo: true
#| eval: false
#| code-line-numbers: "9,10"

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat,
                   group = group,
                   fill = region),
               color = "black") +
  theme_minimal() +
  guides(fill = "none") +
  theme(panel.background =
          element_rect(fill = "lightblue"))
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot3
#| echo: false
#| warning: false
#| fig-width: 7
#| fig-height: 5
```
:::
:::

---

## Change the Map Projection

The earth is not flat, but we visualize it in 2D — there are many projections to choose from.  
See [{mapproj}](https://rdrr.io/cran/mapproj/man/mapproject.html) for available options.

[Use a Mercator projection:]{.orange}

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot4
#| echo: true
#| eval: false
#| code-line-numbers: "11,12"

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat,
                   group = group,
                   fill = region),
               color = "black") +
  theme_minimal() +
  guides(fill = "none") +
  theme(panel.background =
          element_rect(fill = "lightblue")) +
  coord_map(projection = "mercator",
            xlim = c(-180, 180))
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot4
#| echo: false
#| warning: false
#| fig-width: 7
#| fig-height: 5
```
:::
:::

---

## Change the Map Projection

[Use a sinusoidal projection:]{.orange}

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot5
#| echo: true
#| eval: false
#| code-line-numbers: "11,12"

ggplot() +
  geom_polygon(data = world,
               aes(x = long, y = lat,
                   group = group,
                   fill = region),
               color = "black") +
  theme_minimal() +
  guides(fill = "none") +
  theme(panel.background =
          element_rect(fill = "lightblue")) +
  coord_map(projection = "sinusoidal",
            xlim = c(-180, 180))
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot5
#| echo: false
#| warning: false
#| fig-width: 7
#| fig-height: 5
```
:::
:::

---

## Map of Just California

Filter the [states]{.var} dataset to California:

```{r}
#| echo: true

CA_data <- states |>
  filter(region == "california")
```

::: {.callout-tip}
**Think, pair, share:** Make a simple map of California using what you just learned. No peeking at the next slide!
:::

---

## Map of California

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot6
#| echo: true
#| eval: false
#| code-line-numbers: "7"

ggplot() +
  geom_polygon(data = CA_data,
               aes(x = long, y = lat,
                   group = group),
               color = "black") +
  coord_map() +
  theme_void()
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot6
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 7
```
:::
:::

---

## Adding Multiple Layers of Data

[Plot the population of every county in California.]{.orange}

First, we need county-level polygons and join them to our population dataset.

```{r}
#| echo: true

head(counties)[1:3, ]
```

```{r}
#| echo: true

head(popdata)
```

---

## Wrangle the Data

[Make column names match before joining.]{.orange}

```{r}
#| echo: true
#| message: false

CApop_county <- popdata |>
  select("subregion" = County, Population) |>   # rename to match counties
  inner_join(counties) |>
  filter(region == "california")                 # some county names repeat in other states
```

```{r}
#| echo: true

head(CApop_county)
```

---

## Map of CA Population by County

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot7
#| echo: true
#| eval: false


ggplot() +
  geom_polygon(data = CApop_county,
               aes(x = long, y = lat,
                   group = group,
                   fill = Population),
               color = "black") +
  coord_map() +
  theme_void()
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot7
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 7
```
:::
:::

---

## Log Scale for Easier Interpretation

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot8
#| echo: true
#| eval: false
#| code-line-numbers: "9"

ggplot() +
  geom_polygon(data = CApop_county,
               aes(x = long, y = lat,
                   group = group,
                   fill = Population),
               color = "black") +
  coord_map() +
  theme_void() +
  scale_fill_gradient(transform = "log10")
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot8
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 7
```
:::
:::

::: {.callout-tip}
[transform = "log10"]{.param} compresses the color scale so that less-populated counties are still visible alongside Los Angeles.
:::

---

## Add a Layer of Points

The [stars]{.var} dataset has seastar counts (per m²) at field sites in California.

```{r}
#| echo: true

head(stars)
```

---

## Overlay Seastar Sites

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot9
#| echo: true
#| eval: false
#| code-line-numbers: "7,8,9"

ggplot() +
  geom_polygon(data = CApop_county,
               aes(x = long, y = lat,
                   group = group,
                   fill = Population),
               color = "black") +
  geom_point(data = stars,
             aes(x = long,
                 y = lat)) +
  coord_map() +
  theme_void() +
  scale_fill_gradient(transform = "log10")
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot9
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 7
```
:::
:::

---

## Size Points by Seastar Count

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot10
#| echo: true
#| eval: false
#| code-line-numbers: "9"

ggplot() +
  geom_polygon(data = CApop_county,
               aes(x = long, y = lat,
                   group = group,
                   fill = Population),
               color = "black") +
  geom_point(data = stars,
             aes(x = long, y = lat,
                 size = star_no)) +
  coord_map() +
  theme_void() +
  scale_fill_gradient(transform = "log10")
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot10
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 7
```
:::
:::

---

## Better Legend Labels

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: plot11
#| echo: true
#| eval: false
#| code-line-numbers: "13"

ggplot() +
  geom_polygon(data = CApop_county,
               aes(x = long, y = lat,
                   group = group,
                   fill = Population),
               color = "black") +
  geom_point(data = stars,
             aes(x = long, y = lat,
                 size = star_no)) +
  coord_map() +
  theme_void() +
  scale_fill_gradient(transform = "log10") +
  labs(size = "# stars/m²")
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: plot11
#| echo: false
#| warning: false
#| fig-width: 5
#| fig-height: 7
```
:::
:::

---

## Save the Map

```{r}
#| label: plot12
#| echo: true
#| eval: false
#| code-line-numbers: "15"

ggplot() +
  geom_polygon(data = CApop_county,
               aes(x = long, y = lat,
                   group = group,
                   fill = Population),
               color = "black") +
  geom_point(data = stars,
             aes(x = long, y = lat,
                 size = star_no)) +
  coord_map() +
  theme_void() +
  scale_fill_gradient(transform = "log10") +
  labs(size = "# stars/m²")

ggsave(here("Week_07", "output", "CApop.pdf"))
```

::: {.callout-tip}
[ggsave()]{.fn} saves the **last** plot by default. Run it on the line immediately after your ggplot code, or pass the plot object explicitly.
:::

---

## The Modern Approach: {sf} + {rnaturalearth} {.inverse}

Everything we just did works — but it's the old way.  
Here is a quick preview of where the field has moved.

---

## Why {sf}?

The [{sf}]{.fn} ("simple features") package is the current standard for spatial data in R.

::: {.incremental}
- Spatial objects are just **data frames** — all tidyverse verbs work on them
- Handles projections, CRS transformations, spatial joins natively
- [geom_sf()]{.fn} replaces [geom_polygon()]{.fn} — no need to manage [group]{.var} manually
- Compatible with a rich ecosystem: [{tigris}]{.fn}, [{rnaturalearth}]{.fn}, [{maptiles}]{.fn}, [{leaflet}]{.fn}
:::

::: {.callout-tip}
For any serious spatial analysis — buffering, intersections, area calculations — [{sf}]{.fn} is the tool you want. Today's [{maps}]{.fn} approach is great for quick plots, but it won't scale to real GIS work.
:::

---

## {rnaturalearth} — World Map with geom_sf()

Install once:

```{r}
#| echo: true
#| eval: false

install.packages(c("sf", "rnaturalearth", "rnaturalearthdata"))
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| output-location: column
#| code-line-numbers: "5,6,9,10"
#| fig-width: 7
#| fig-height: 5

library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

world_sf <- ne_countries(scale = "medium",
                         returnclass = "sf")

ggplot(world_sf) +
  geom_sf(fill = "lightgray",
          color = "white") +
  theme_void()
```

No [group]{.var}! No longitude/latitude columns to manage! [{sf}]{.fn} handles all of that.

---

## Projections with coord_sf()

[`coord_sf()`]{.fn} handles projections using standard **EPSG codes** — much more rigorous than [{mapproj}]{.fn}:

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: sf_proj
#| echo: true
#| eval: false
#| code-line-numbers: "3"

ggplot(world_sf) +
  geom_sf(fill = "lightgray", color = "white") +
  coord_sf(crs = "+proj=robin") + # Robinson projection
  theme_void() +
  labs(title = "Robinson Projection")
```
:::
::: {.column width="50%"}
```{r}
#| ref.label: sf_proj
#| echo: false
#| warning: false
#| fig-width: 7
#| fig-height: 5
```
:::
:::

::: {.callout-note}
Common CRS choices: [`"+proj=robin"`]{.param} (Robinson), [`4326`]{.param} (WGS84 / GPS), [`3857`]{.param} (Web Mercator).  
Learn more at [map chapter in ggplot2 book](https://ggplot2-book.org/maps#sec-sf).
:::

---

## Total Awesome R Package — {ggdogs} 🐾

```{r}
#| echo: true
#| eval: false

remotes::install_github("R-CoderDotCom/ggdogs@main")
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 4

library(ggdogs)
ggplot(mtcars) +
  geom_dog(aes(mpg, wt), dog = "pug", size = 5)
```

---

## Homework

Make a map **using Quarto** with one of the following TidyTuesday datasets. Save your script and output appropriately.

- [Plastic Pollution](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-01-26/readme.md)
- [Cats on the Move!](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-01-31/readme.md)
- [Where Is Bigfoot?](https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-09-13)
- [Meteorites](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-06-11)

::: {.callout-tip}
The map does not have to be complex — a clean, labeled [geom_polygon()]{.fn} or [geom_sf()]{.fn} map with your data overlaid is exactly right.
:::

---

## Thanks! {.inverse .center}

Slides created with [Quarto](https://quarto.org) and [RevealJS](https://revealjs.com/).  
Some content adapted from [Making Maps with R](https://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html/) and [Geocomputation with R](https://geocompr.robinlovelace.net/).
