---
title: "Computer Modeling (OCN 682)"
subtitle: "Intro to Maps Part 2"
author: "Dr. Nyssa Silbiger"
institute: "UH Fall 2026"
date: today
resources:
  - custom-theme.scss
format:
  revealjs:
    theme: [default, custom-theme.scss]
    highlight-style: github
    code-line-numbers: false
    width: 1500
    height: 1000
    preview-links: auto
    self-contained-math: true
    embed-resources: false
    slide-number: c/t
    footer: "OCN 682 | Week 7 | Intro to Maps Part 2"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)
```

## Outline 

Working with Maps — Part 2

::: {.incremental}
1. Review
1. Tile and satellite basemaps with [{maptiles}]{.fn} 
1. Overlaying data + annotations with [{ggspatial}]{.fn}
1. Interactive maps with [{leaflet}]{.fn}
1. Geocoding with [{tidygeocoder}]{.fn}
:::

**References:**
- [maptiles documentation](https://github.com/riatelab/maptiles)
- [tidyterra documentation](https://dieghernan.github.io/tidyterra/)
- [Leaflet for R](https://rstudio.github.io/leaflet/)

---

## Review

::: {.callout-warning}
**If your map looks like a tangle of crazy lines, what did you most likely forget?**
:::

. . .

[`group = group`]{.param} in your [geom_polygon()]{.fn} aesthetic.

Without it, ggplot connects every vertex in one continuous path — it has no idea where one polygon ends and the next begins.

---

## Install Today's Packages

```{r}
#| echo: true
#| eval: false

install.packages(c(

  "maptiles",      # download tile basemaps (satellite, street, terrain)
  "tidyterra",     # plot terra rasters in ggplot2
  "leaflet",       # interactive HTML maps
  "tidygeocoder",  # free geocoding (OpenStreetMap / Census)
  "ggspatial"      # scale bars and north arrows
))
```

::: {.callout-tip}
None of these require an API key or a credit card.
:::

---

## Load Libraries

```{r}
#| echo: true
#| message: false
#| warning: false

library(tidyverse)
library(here)
library(sf)
library(maptiles)
library(tidyterra)
library(leaflet)
library(tidygeocoder)
library(ggspatial)
```

---

## Data for Today

```{r}
#| echo: true
#| message: false
#| warning: false

ChemData <- read_csv(here("Week_07", "data", "chemicaldata_maunalua.csv"))

glimpse(ChemData)
```

---

## Getting Tile Basemaps with {maptiles} {.inverse}

[{maptiles}]{.fn} downloads satellite and street map tiles  
from free public providers directly into R.

---

## What Is a Map Tile?

Online maps (Google, OpenStreetMap, ESRI) are not single images — they are made of thousands of small **256 × 256 pixel tiles** stitched together.

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- Tiles are organized by **zoom level** and **grid position**
- At zoom 1, a handful of tiles cover the whole world
- At zoom 17, thousands of tiles cover a single neighborhood
- [{maptiles}]{.fn} figures out which tiles you need, downloads them, and stitches them into one image
:::
:::
::: {.column width="45%"}
::: {.callout-tip}
This is exactly how Google Maps works — [{maptiles}]{.fn} just accesses the same tiles directly, using providers that allow free access.
:::
:::
:::

---

## Step 1 — Define Your Area of Interest

[get_tiles()]{.fn} needs to know **what area to download**. We give it a **bounding box** — a rectangle defined by its corner coordinates.

```{r}
#| echo: true
#| message: false
#| warning: false

# Define a bounding box around Wailupe, Oahu
# xmin/xmax = westmost/eastmost longitude
# ymin/ymax = southernmost/northernmost latitude
wailupe_bbox <- st_as_sfc(
  st_bbox(c(xmin = -157.772, ymin = 21.267,
            xmax = -157.752, ymax = 21.281),
          crs = 4326) # crs = 4326 means WGS84 — the lat/lon system your GPS uses
)
```

::: {.callout-note}
**[st_bbox()]{.fn}** creates the bounding box rectangle.  
**[st_as_sfc()]{.fn}** converts it to an [{sf}]{.fn} geometry object so [{maptiles}]{.fn} can read it.  
**[crs = 4326]{.param}** specifies WGS84 — almost all GPS and web map data uses this coordinate system.
:::

---

## Step 2 — Download the Tiles

```{r}
#| echo: true
#| message: false
#| warning: false

tiles_satellite <- get_tiles(
  wailupe_bbox,               # the area to download
  provider = "Esri.WorldImagery", # which tile source to use
  zoom = 16,                  # level of detail (3 = continent, 20 = building)
  crop = TRUE,                # clip to our exact bbox (vs. slightly larger)
  cachedir = here("Week_07", "data", "tiles_cache") # save tiles to disk
)
```

::: {.callout-tip}
**[cachedir]{.param}** saves the raw tile images to disk. The next time you render this document, [{maptiles}]{.fn} will read from disk instead of re-downloading. Always include this when building presentations.
:::

---

## What Does get_tiles() Return?

```{r}
#| echo: true

class(tiles_satellite)
```

. . .

A **[SpatRaster]{.var}** — a raster image (grid of pixels) from the [{terra}]{.fn} package.

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- A raster is just a matrix of pixel values — like a photo
- [tiles_satellite]{.var} has **3 layers**: red, green, and blue
- We can't use `geom_raster()` because that expects a data frame, not a terra object
- [{tidyterra}]{.fn} bridges this gap with [geom_spatraster_rgb()]{.fn}
:::
:::
::: {.column width="45%"}
```{r}
#| echo: true

terra::nlyr(tiles_satellite) # number of layers (R, G, B)
```
:::
:::

---

## Step 3 — Plot with geom_spatraster_rgb()

```{r}
#| echo: true
#| output-location: column
#| fig-width: 6
#| fig-height: 5

# geom_spatraster_rgb() is from {tidyterra}
# It reads the 3 color layers and renders them as a photo
ggplot() +
  geom_spatraster_rgb(data = tiles_satellite) +
  coord_sf(crs = 4326) +
  theme_minimal()
```

::: {.callout-note}
[coord_sf(crs = 4326)]{.fn} tells ggplot to treat your axes as geographic coordinates in WGS84. Always include it when plotting spatial data — it ensures the map is not distorted and that scale bars will work correctly later.
:::

---

## Zoom Levels

The [zoom]{.param} argument controls how much detail is shown. Larger number = more detail but more tiles to download.

::: {.columns}
::: {.column width="33%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 4
#| fig-height: 4

tiles_z10 <- get_tiles(wailupe_bbox, provider = "Esri.WorldImagery", zoom = 10, crop = TRUE, cachedir = here("Week_07", "data", "tiles_cache"))
ggplot() +
  geom_spatraster_rgb(data = tiles_z10) +
  coord_sf(crs = 4326) +
  theme_void() +
  labs(title = "zoom = 10 (island)")
```
:::
::: {.column width="33%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 4
#| fig-height: 4

tiles_z14 <- get_tiles(wailupe_bbox, provider = "Esri.WorldImagery", zoom = 14, crop = TRUE, cachedir = here("Week_07", "data", "tiles_cache"))
ggplot() +
  geom_spatraster_rgb(data = tiles_z14) +
  coord_sf(crs = 4326) +
  theme_void() +
  labs(title = "zoom = 14 (neighborhood)")
```
:::
::: {.column width="33%"}
```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 4
#| fig-height: 4

tiles_z17 <- get_tiles(wailupe_bbox, provider = "Esri.WorldImagery", zoom = 17, crop = TRUE, cachedir = here("Week_07", "data", "tiles_cache"))
ggplot() +
  geom_spatraster_rgb(data = tiles_z17) +
  coord_sf(crs = 4326) +
  theme_void() +
  labs(title = "zoom = 17 (site level)")
```
:::
:::

---

## Different Tile Providers

Just change the [provider]{.param} string — all are free with no API key:

::: {.columns}
::: {.column width="50%"}

| Provider string | Style |
|---|---|
| `"Esri.WorldImagery"` | Satellite |
| `"OpenStreetMap"` | Street map |
| `"CartoDB.Positron"` | Minimal/clean |
| `"Esri.WorldStreetMap"` | ESRI street |
| `"Esri.WorldTerrain"` | Shaded terrain |

:::
::: {.column width="50%"}
```{r}
#| echo: true
#| eval: false

tiles_street <- get_tiles(
  wailupe_bbox,
  provider = "CartoDB.Positron",
  zoom = 14,
  cachedir = here("Week_07", "data", "tiles_cache")
)
```
:::
:::

::: {.callout-note}
Browse all available providers at [leaflet-extras.github.io/leaflet-providers](https://leaflet-extras.github.io/leaflet-providers/preview/).
:::

---

## Deriving the Bbox from Your Data

Instead of hard-coding coordinates, generate the bounding box directly from your data:

```{r}
#| echo: true
#| message: false
#| warning: false

# Step 1: convert your data frame to an sf object
# coords tells sf which columns hold longitude and latitude
ChemData_sf <- st_as_sf(ChemData,
                         coords = c("Long", "Lat"),
                         crs = 4326)

# Step 2: get the bounding box and add a small buffer
# the + c(...) pads all four sides by ~0.4 km so points aren't at the edge
bbox_chem <- st_bbox(ChemData_sf) + c(-0.004, -0.004, 0.004, 0.004)

# Step 3: download tiles for that extent
Map_satellite <- get_tiles(
  st_as_sfc(bbox_chem),
  provider = "Esri.WorldImagery",
  zoom = 15,
  crop = TRUE,
  cachedir = here("Week_07", "data", "tiles_cache")
)
```

---

## Overlay Point Data

Once you have a tile basemap, add ggplot layers on top exactly as you normally would:

```{r}
#| echo: true
#| output-location: column
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 5

ggplot() +
  geom_spatraster_rgb(data = Map_satellite) +
  geom_point(data = ChemData,
             aes(x = Long, y = Lat,
                 color = Salinity),
             size = 4) +
  scale_color_viridis_c() +
  coord_sf(crs = 4326) +
  theme_minimal()
```

::: {.callout-tip}
Any [geom_*()]{.fn} layer works on top of [geom_spatraster_rgb()]{.fn} — points, lines, polygons. The tile is just another ggplot layer.
:::

---

## Add a Scale Bar and North Arrow

[{ggspatial}]{.fn} provides two annotation functions that are spatial-aware — they calculate real-world distances and magnetic north automatically.

```{r}
#| echo: true
#| output-location: column
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 5

ggplot() +
  geom_spatraster_rgb(data = Map_satellite) +
  geom_point(data = ChemData,
             aes(x = Long, y = Lat,
                 color = Salinity),
             size = 4) +
  scale_color_viridis_c() +
  annotation_scale(           # scale bar
    location = "bl",          # bl = bottom-left
    bar_cols = c("yellow", "white")) +
  annotation_north_arrow(     # north arrow
    location = "tl") +        # tl = top-left
  coord_sf(crs = 4326) +
  theme_minimal()
```

---

## Polish and Save

```{r}
#| echo: true
#| eval: false

ggplot() +
  geom_spatraster_rgb(data = Map_satellite) +
  geom_point(data = ChemData,
             aes(x = Long, y = Lat, color = Salinity),
             size = 4) +
  scale_color_viridis_c(name = "Salinity (ppt)") +
  annotation_scale(location = "bl",
                   bar_cols = c("yellow", "white")) +
  annotation_north_arrow(location = "tl") +
  coord_sf(crs = 4326) +
  labs(title = "Maunalua Bay — Water Chemistry",
       caption = "Basemap: ESRI World Imagery") +
  theme_minimal()

ggsave(here("Week_07", "output", "maunalua_map.pdf"), width = 8, height = 6)
```

---

## Interactive Maps with {leaflet} {.inverse}

Static maps are great for publications.  
Interactive maps let you **zoom, pan, and click on your data** — and embed directly in HTML documents.

---

## How Leaflet Works

[{leaflet}]{.fn} builds maps by **piping layers together with `|>`** — similar in concept to ggplot's `+`, but using pipes instead.

```{r}
#| echo: true
#| eval: false

leaflet() |>                       # 1. create the map widget
  addProviderTiles(...) |>         # 2. add a basemap tile layer
  addCircleMarkers(...) |>         # 3. add your data as markers
  addLegend(...)                   # 4. add a color legend
```

::: {.incremental}
- [leaflet()]{.fn} — initializes an empty map widget. Pass your data frame here.
- [addProviderTiles()]{.fn} — adds a basemap. Uses `providers$Name` to select from hundreds of free tile sources.
- [add*()]{.fn} functions — each one adds a new layer (markers, polygons, popups, legends...)
- The result is an **interactive HTML widget** — it works in RStudio, Quarto HTML, and any web browser.
:::

---

## Your First Leaflet Map

```{r}
#| echo: true
#| eval: true

leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  setView(lng = -157.762, lat = 21.274, zoom = 14)
```

::: {.callout-note}
[setView()]{.fn} centers the map on a specific location. [lng]{.param} and [lat]{.param} are the coordinates, and [zoom]{.param} works exactly like in [{maptiles}]{.fn}.
:::

---

## The ~ Formula Notation

In leaflet, column names are written with a `~` prefix:

```{r}
#| echo: true
#| eval: false

addCircleMarkers(lng = ~Long, lat = ~Lat, color = ~pal(Salinity))
```

::: {.incremental}
- `~Long` means *"use the [Long]{.var} column from the data frame"*
- It is leaflet's equivalent of `aes(x = Long)` in ggplot
- You will see `~` anywhere leaflet needs to reference a column
:::

::: {.callout-tip}
The `~` is leaflet's way of saying "evaluate this as a formula against the data." Without it, leaflet treats the argument as a fixed value, not a column reference.
:::

---

## Add a Color Palette

Leaflet needs a **color function** that converts numeric values to hex colors. [colorNumeric()]{.fn} creates one:

```{r}
#| echo: true

# colorNumeric() returns a FUNCTION, not a color vector
pal <- colorNumeric(
  palette = "viridis",            # any RColorBrewer or viridis palette name
  domain = ChemData$Salinity      # the full range of values to map
)

# Use it by calling pal(value) — it returns a hex color
pal(30)  # what color does salinity = 30 get?
```

. . .

```{r}
#| echo: true

pal(40)  # higher salinity → different color
```

---

## Add Data Points with Color

```{r}
#| echo: true

pal <- colorNumeric("viridis", domain = ChemData$Salinity)

leaflet(ChemData) |>           # pass data frame here
  addProviderTiles(providers$Esri.WorldImagery) |>
  addCircleMarkers(
    lng = ~Long, lat = ~Lat,   # which columns are coordinates
    color = ~pal(Salinity),    # apply our color function to Salinity
    radius = 6,
    fillOpacity = 0.9,
    popup = ~paste("Salinity:", Salinity, "ppt",  # click a point to see this
                   "<br>Site:", Site)) |>
  addLegend(
    pal = pal,                 # the same color function
    values = ~Salinity,        # which column drives the legend
    title = "Salinity (ppt)",
    position = "bottomright")
```

---

## popup vs. label

::: {.columns}
::: {.column width="50%"}

**[popup]{.param}** — appears when you **click** a marker

```{r}
#| echo: true
#| eval: false

addCircleMarkers(
  popup = ~paste("Site:", Site,
                 "<br>Salinity:", Salinity))
```

HTML tags work inside popups (`<br>` = line break).

:::
::: {.column width="50%"}

**[label]{.param}** — appears when you **hover** over a marker

```{r}
#| echo: true
#| eval: false

addCircleMarkers(
  label = ~Site)
```

Labels are plain text only (no HTML).

:::
:::

::: {.callout-tip}
Use [popup]{.param} for detailed information (multiple fields, formatted text).  
Use [label]{.param} for a quick identifier on hover.
:::

---

## Geocoding with {tidygeocoder} {.inverse}

**Geocoding** = converting a place name or address into latitude and longitude coordinates.

---

## What Is Geocoding?

::: {.columns}
::: {.column width="55%"}
::: {.incremental}
- You know *where* you want to map, but not the exact coordinates
- Geocoding services look up addresses and place names in a database and return lat/lon
- [{tidygeocoder}]{.fn} does this for free using:
  - **OpenStreetMap** (`method = "osm"`) — best for landmarks and place names worldwide
  - **US Census** (`method = "census"`) — best for US street addresses
:::
:::
::: {.column width="45%"}
::: {.callout-warning}
Geocoding requires an **internet connection** and results depend on the service's database. OSM is very good for well-known places but may fail on obscure locations or unusual search strings.
:::
:::
:::

---

## geo() — Convert a Place Name to Coordinates

[geo()]{.fn} geocodes a single location string and returns a one-row tibble:

```{r}
#| echo: true
#| message: false
#| warning: false

geo("University of Hawaii at Manoa", method = "osm")
```

```{r}
#| echo: true
#| message: false
#| warning: false

geo("Pearl Harbor, Oahu, Hawaii", method = "osm")
```

::: {.callout-tip}
If a location returns [NA]{.var}, try a more specific search string — add the city, state, or country. OSM often needs more context for landmarks that share names across locations.
:::

---

## geocode() — Geocode a Data Frame

[geocode()]{.fn} (note: no "r") operates on a data frame and adds lat/lon columns:

```{r}
#| echo: true
#| message: false
#| warning: false

oahu_sites <- tribble(
  ~place,
  "University of Hawaii at Manoa",
  "Honolulu International Airport",
  "Pearl Harbor National Memorial, Hawaii",
  "Diamond Head State Monument, Hawaii"
)

oahu_geocoded <- oahu_sites |>
  geocode(place, method = "osm") # geocode the 'place' column

oahu_geocoded
```

---

## Map Geocoded Locations

```{r}
#| echo: true
#| message: false
#| warning: false

leaflet(oahu_geocoded) |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  addMarkers(
    lng = ~long, lat = ~lat,   # tidygeocoder always returns 'lat' and 'long'
    label = ~place)            # hover to see place name
```

---

## Totally Awesome R Package — {emojifont}

```{r}
#| echo: true
#| eval: false

install.packages("emojifont")
```

```{r}
#| echo: true
#| message: false
#| warning: false
#| output-location: column
#| fig-width: 5
#| fig-height: 4

library(emojifont)

search_emoji("smile")
```

```{r}
#| echo: true
#| output-location: column
#| fig-width: 5
#| fig-height: 4

ggplot() +
  geom_emoji("smile_cat",
             x = 1:5, y = 1:5,
             size = 10)
```

[More info: emojifont vignette](https://cran.r-project.org/web/packages/emojifont/vignettes/emojifont.html)

---

## Summary: Static vs. Interactive Maps

| | Static [{maptiles}]{.fn} + [{tidyterra}]{.fn} | Interactive [{leaflet}]{.fn} |
|---|---|---|
| **Output** | PDF, PNG, Quarto figure | HTML widget |
| **Best for** | Publications, reports | Exploration, sharing online |
| **Layers** | Any [geom_*()]{.fn} via ggplot | [add*()]{.fn} functions via pipe |
| **Color palettes** | `scale_color_*()` | [colorNumeric()]{.fn} etc. |
| **Scale bars** | [{ggspatial}]{.fn} | Built-in scale control |


---

## Thanks! {.inverse .center}

Slides created with [Quarto](https://quarto.org) and [RevealJS](https://revealjs.com/).  
Some content adapted from [Making Maps with R](http://stat405.had.co.nz/ggmap.pdf).

**Tile data © [ESRI](https://www.esri.com) | © [OpenStreetMap contributors](https://www.openstreetmap.org/copyright)**
