---
title: "Computer Modeling (OCN 682)"
subtitle: "Data wrangling: joins & dates with lubridate"
author: "Dr. Nyssa Silbiger"
institute: "UH Fall 2026"
date: today
resources:
  - custom-theme.scss
format:
  revealjs:
    theme: [default, custom-theme.scss]
    highlight-style: github
    code-line-numbers: true
    width: 1500
    height: 1000
    preview-links: auto
    self-contained-math: true
    embed-resources: false
    slide-number: c/t
    footer: "OCN 682 | Week 5 | Joins & Dates with lubridate"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)
library(here)
library(lubridate)
```

## Outline of class

::: {.incremental}
1. Review of tidyr
2. Learning joins (part of the [dplyr]{.fn} package)
   - [left_join()]{.fn} — with real data
   - Creating tibbles
   - [right_join()]{.fn}, [inner_join()]{.fn}, [full_join()]{.fn}
   - [semi_join()]{.fn}, [anti_join()]{.fn}
3. Dates and times using [lubridate]{.fn}
   - Converting dates and times
   - Extracting date/time elements
   - Manipulating and rounding dates
:::

**Homework:** Practice with joins and dates in online assignment

---

## Review {.center}

::: {.incremental}
1. What function do you use to split up one column into multiple columns?
2. What function do you use to transition a wide data frame into a long data frame?
:::

---

# Part 1: Data Joins

---

## Intro to different types of joins

Let's say you have multiple hierarchies of data and you want to join them into one big data set. Join functions are the way you would do it.

::: {.columns}

::: {.column width="50%"}
### Data 1

| Site_number | Treatment | Nitrate | Temp |
|---|---|---|---|
| 1 | High | 1.2 | 7.2 |
| 2 | High | 3.0 | 7.8 |
| 3 | Low | 2.4 | 8.0 |
| 4 | Low | 5.1 | 8.0 |
| 5 | Low | 1.1 | 7.9 |
:::

::: {.column width="50%"}
### Data 2

| Site_Number | Sample_number | Biomass |
|---|---|---|
| 1 | 1 | 12.2 |
| 1 | 2 | 14.0 |
| 1 | 3 | 11.9 |
| 1 | 4 | 10.0 |
:::

:::

---

## Identifying the key

Which column is consistent across both datasets?

::: {.columns}

::: {.column width="50%"}
### Data 1

| Site_Number | Treatment | Nitrate | Temp |
|---|---|---|---|
| [1]{.orange} | [High]{.orange} | [1.2]{.orange} | [7.2]{.orange} |
| 2 | High | 3.0 | 7.8 |
| 3 | Low | 2.4 | 8.0 |
| 4 | Low | 5.1 | 8.0 |
| 5 | Low | 1.1 | 7.9 |
:::

::: {.column width="50%"}
### Data 2

| Site_Number | Sample_number | Biomass |
|---|---|---|
| [1]{.orange} | [1]{.orange} | [12.2]{.orange} |
| [1]{.orange} | [2]{.orange} | [14.0]{.orange} |
| [1]{.orange} | [3]{.orange} | [11.9]{.orange} |
| [1]{.orange} | [4]{.orange} | [10.0]{.orange} |
:::

:::

::: {.fragment}
**The column name must be identical** in both dataframes (spelling, capitalization, everything). This is your **key**.

Whenever collecting data, include a unique ID for every sample. This makes joining datasets much easier!
:::

---

## Real-world example

From **Becker and Silbiger (2020) Journal of Experimental Biology**.

Danielle tested how nutrient loading and sedimentation affected thermal performance metrics of corals in Mo'orea, French Polynesia. She collected 10 corals from six sites with a wide range in nutrients/sedimentation.

**Three datasets:**
1. `site.characteristics.data.csv` — Environmental data from each site
2. `Topt_data.csv` — Thermal performance curves
3. `Data_dictionary.csv` — Column descriptions

---

## Creating sample tibbles

Let's create two small example datasets:

```{r}
#| echo: true
T1 <- tibble(
  Site.ID = c("A", "B", "C", "D"),
  Temperature = c(14.1, 16.7, 15.3, 12.8)
)

T1
```

---

## Creating sample tibbles (continued)

Now create a second tibble:

```{r}
#| echo: true
T2 <- tibble(
  Site.ID = c("A", "B", "D", "E"),
  pH = c(7.3, 7.8, 8.1, 7.9)
)

T2
```

---

## [left_join()]{.fn}

[left_join()]{.fn} keeps all rows from the left (first) dataframe and adds matching rows from the right:

```{r}
#| echo: true
left_join(T1, T2)
```

Notice that Site "C" stays (from T1) but "E" is excluded (only in T2).

---

## [right_join()]{.fn}

[right_join()]{.fn} keeps all rows from the right (second) dataframe:

```{r}
#| echo: true
right_join(T1, T2)
```

Now Site "E" is included, but "C" is excluded.

---

## [inner_join()]{.fn}

[inner_join()]{.fn} keeps **only** rows that exist in both dataframes:

```{r}
#| echo: true
inner_join(T1, T2)
```

Only sites A, B, and D are kept. C and E are dropped.

::: {.callout-warning}
## Data loss risk
[inner_join()]{.fn} drops unmatched rows. Always check your result to ensure you haven't unexpectedly lost data.
:::

---

## [full_join()]{.fn}

[full_join()]{.fn} keeps **all** rows from both dataframes:

```{r}
#| echo: true
full_join(T1, T2)
```

All sites (A, B, C, D, E) are kept, with [NA]{.err} filling in missing values.

---

## [semi_join()]{.fn}

[semi_join()]{.fn} keeps rows from the first dataframe where there are matches in the second, but only returns columns from the first:

```{r}
#| echo: true
semi_join(T1, T2)
```

Only sites A, B, D from T1, and only the Temperature column (from T1).

---

## [anti_join()]{.fn}

[anti_join()]{.fn} returns rows in the first dataframe that **do not** match the second:

```{r}
#| echo: true
anti_join(T1, T2)
```

Only Site C from T1 is returned (it has no match in T2).

::: {.callout-tip}
## Finding missing data
Use [anti_join()]{.fn} to identify unmatched records and detect data quality issues.
:::

---

## Summary of join types

| Join | Keeps | Use when |
|---|---|---|
| [left_join()]{.fn} | All left rows + matches | Primary dataset is on the left |
| [right_join()]{.fn} | All right rows + matches | Primary dataset is on the right |
| [inner_join()]{.fn} | Only complete matches | Need only complete pairs |
| [full_join()]{.fn} | All rows from both | Want to keep everything |
| [semi_join()]{.fn} | Left rows with matches | Filtering without duplication |
| [anti_join()]{.fn} | Left rows without matches | Finding unmatched data |



---

# Part 2: Dates & Times with lubridate

---

## What time is it now?

```{r}
#| echo: true
now()
```

---

## Current time in different timezones

```{r}
#| echo: true
now(tzone = "EST")  # East Coast
now(tzone = "GMT")  # Greenwich Mean Time
```

---

## Just the date

```{r}
#| echo: true
today()
```

---

## Just the date in a specific timezone

```{r}
#| echo: true
today(tzone = "GMT")
```

---

## Time checks

```{r}
#| echo: true
am(now())        # Is it morning?
leap_year(now()) # Is it a leap year?
```

---

## Date specifications

[lubridate]{.fn} guesses date formats from character strings using **y**ear, **m**onth, **d**ay abbreviations:

| Date format | Function |
|---|---|
| 2021-02-24 | [ymd()]{.fn} |
| 02/24/2021 | [mdy()]{.fn} |
| February 24 2021 | [mdy()]{.fn} |
| 24/02/2021 | [dmy()]{.fn} |

::: {.callout-warning}
## Dates must be characters
Always ensure your date column is a [character]{.dt}, not a factor. Use `as.character()` if needed.
:::

---

## Convert dates: ISO format (YYYY-MM-DD)

```{r}
#| echo: true
ymd("2021-02-24")
```

---

## Convert dates: US format (MM/DD/YYYY)

```{r}
#| echo: true
mdy("02/24/2021")
```

---

## Convert dates: Written month

```{r}
#| echo: true
mdy("February 24 2021")
```

---

## Convert dates: European format (DD/MM/YYYY)

```{r}
#| echo: true
dmy("24/02/2021")
```

---

## Date & time specifications

| Format | Function |
|---|---|
| 2021-02-24 10:22:20 PM | [ymd_hms()]{.fn} |
| 02/24/2021 22:22:20 | [mdy_hms()]{.fn} |
| February 24 2021 10:22 PM | [mdy_hm()]{.fn} |

---

## Convert date and time: ISO format with time

```{r}
#| echo: true
ymd_hms("2021-02-24 10:22:20 PM")
```

---

## Convert date and time: US format with seconds

```{r}
#| echo: true
mdy_hms("02/24/2021 22:22:20")
```

---

## Convert date and time: Written month with hour/minute

```{r}
#| echo: true
mdy_hm("February 24 2021 10:22 PM")
```

---

## Create a vector of datetimes

```{r}
#| echo: true
datetimes <- c(
  "02/24/2021 22:22:20",
  "02/25/2021 11:21:10",
  "02/26/2021 8:01:52"
)

datetimes
```

---

## Convert the vector to datetime objects

```{r}
#| echo: true
datetimes <- mdy_hms(datetimes)

datetimes
```

Notice the output now shows [POSIXct]{.dt} dates with times!

---

## Extract the month (as number)

```{r}
#| echo: true
month(datetimes)
```

---

## Extract the month (as abbreviated label)

```{r}
#| echo: true
month(datetimes, label = TRUE)
```

---

## Extract the month (as full name)

```{r}
#| echo: true
month(datetimes, label = TRUE, abbr = FALSE)
```

---

## Extract the day of the month

```{r}
#| echo: true
day(datetimes)
```

---

## Extract the day of the week

```{r}
#| echo: true
wday(datetimes, label = TRUE)
```

---

## Extract hour, minute, and second

```{r}
#| echo: true
hour(datetimes)
minute(datetimes)
second(datetimes)
```

---

## Adding time intervals: Add 4 hours

```{r}
#| echo: true
datetimes + hours(4)
```

::: {.callout-note}
## Singular vs plural
[hour()]{.fn} *extracts* the hour component  
[hours()]{.fn} *adds* hours to a datetime

The same pattern applies to all time units!
:::

---

## Adding time intervals: Add 2 days

```{r}
#| echo: true
datetimes + days(2)
```

You can also add [minutes()]{.fn}, [seconds()]{.fn}, [months()]{.fn}, [years()]{.fn}, and more!

---

## Adding time intervals: Add 1 month

```{r}
#| echo: true
datetimes + months(1)
```

---

## Rounding dates: Round to nearest minute

```{r}
#| echo: true
round_date(datetimes, "minute")
```

---

## Rounding dates: Round to nearest 5 minutes

```{r}
#| echo: true
round_date(datetimes, "5 mins")
```

You can round to any time unit!

---

## Challenge

Read in the conductivity data (`CondData.csv`) and convert the date column to a datetime using the pipe:

---

```{r}
#| echo: true
#| eval: false
CondData <- read_csv(here("Week_05", "data", "CondData.csv")) |>
  mutate(datetime = mdy_hms(datetime_column))
```

::: {.callout-warning}
## Watch for hidden data
Excel may hide seconds or display dates inconsistently. **Always** examine your data after reading it in with [glimpse()]{.fn} or [View()]{.fn}.
:::

---

## Today's awesome R package: [ggcats]{.fn}

Ever wanted to plot with cats in ggplot2?

::: {.columns}

::: {.column width="45%"}

```{r}
#| echo: true
#| eval: false
library(ggplot2)
library(ggcats)

# Create sample data
df <- data.frame(
  x = c(1, 2, 3, 4, 5),
  y = c(2, 4, 3, 5, 4)
)

ggplot(df, aes(x, y)) +
  geom_cat(cat = "nyancat", 
           size = 4) +
  labs(title = "Cats!", 
       x = "X", y = "Y") +
  theme_minimal()
```

:::

::: {.column width="55%"}

```{r}
#| echo: false
#| message: false
#| warning: false
library(ggplot2)
library(ggcats)

df <- data.frame(
  x = c(1, 2, 3, 4, 5),
  y = c(2, 4, 3, 5, 4)
)

ggplot(df, aes(x, y)) +
  geom_cat(cat = "nyancat", size = 4) +
  labs(title = "Plotting with Cats!", x = "X", y = "Y") +
  theme_minimal()
```

:::

:::
---

## Homework {.center}

::: {.incremental}
1. Read in both the conductivity and depth data
2. Convert date columns appropriately
3. Round the conductivity data to the nearest 10 seconds to match depth data
4. Join the two dataframes using [inner_join()]{.fn} (only exact matches)
5. Calculate averages of date, depth, temperature, and salinity **by minute**
6. Make a plot using the averaged data
7. Use pipes throughout (minimize separate dataframes)
8. Add comments to your code!
9. Save output, data, and scripts appropriately
:::

---

## {.center .middle}

# Thanks!

Slides created with [**Quarto**](https://quarto.org/)

References:  
[Tidyexplain GitHub](https://github.com/gadenbuie/tidyexplain) — animated join visualizations  
[lubridate cheatsheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_lubridate.pdf)
